---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

The categories in this data set cover a wide range of crime incidents, from accidents to assaults to homicides to robberies. Some are more minor crimes, such as theft, while others are more serious, such as homicide and sex offenses. These more serious crimes have much lower numbers each month, which is thankfully to be expected. What makes the job of summarizing the data more difficult is that every category save for sex offense has at least one additional, more specific subcategory. For example, the first category in the table is "accident," but just underneath that row is the similar category "accident with impound." If you wanted to write a summary report on accidents per month in Prince George's County, then, you would also have to take into account or maybe even separately analyze the accidents with impounds per month. This process is even more complicated when it comes to categories that have more than one subcategory, such as robbery and assault.

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime <- pg_crime |> mutate(total_incidents = `Jan` + `Feb` + `Mar` + `Apr` + `May` + `Jun` + `Jul` + `Aug` + `Sep` + `Oct` + `Nov` + `Dec`)
pg_crime_averages <- pg_crime |> mutate(average_incidents = (`total_incidents`) / 12)

```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

Accidents, stolen automobiles and theft from auto have the highest average number of incidents per month. Breaking and entering "other" and homicide have the lowest average number of incidents per month, and I would assume that the robbery residential category average would be even lower, although R did not calculate an average for that category because some months had zero residential robbery incidents. This data tells me that most of the crime in Prince George's County involves automobiles in some way, whether it be accidents or theft from or of such vehicles. I'm surprised at how high the stolen auto average is, especially when compared to the subsequent category of "stolen auto & recovered." An average of nearly 400 automobiles are stolen each month, but only about 41 of those stolen vehicles are recovered. Why is it so difficult for the county to track down stolen automobiles? Why are there so many thefts in that category in the first place? What can people do to lower the number of incidents?       


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
pg_crime_averages <- pg_crime_averages |> mutate(june_change = (`Jun` - `average_incidents`)/`average_incidents` * 100)
pg_crime_averages <- pg_crime_averages |> mutate(july_change = (`Jul` - `average_incidents`)/`average_incidents` * 100)
pg_crime_averages <- pg_crime_averages |> mutate(august_change = (`Aug` - `average_incidents`)/`average_incidents` * 100)
summer_crime_averages <- pg_crime_averages

```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

There are some categories where the average number of incidents per month is greater in all three summer months, such as homicides and stolen automobiles and some categories where the average number of incidents per month in greater in 2/3 of the summer months, such as assault shootings and vehicle robberies. But in most cases, there actually seems to be a small decrease in crime in Prince George's County during the months of June, July and August, or at least in 2/3 of those summer months. The data is not very consistent, which leads me to believe that the theory that crime is more likely to occur in the summer months is actually false. Crime incidents might actually be more common during the school year due to UMD's location in the county, although that is another theory that would have to be tested for backup.  

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Stolen automobiles had the second-highest average number of incidents per month in Prince George's County in 2024 compared to all other areas of crime, according to crime data, though not many of those vehicles were found. 

Data shows that while an average of nearly 400 automobiles were stolen each month in 2024, only about 41 were recovered, begging the question of how and why car robbers can so routinely get away with their crimes.  

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
