---
title: "HW7_CarAccidents"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Analyzing Daily Accident Data with Z-Scores

## Introduction

In this analysis, we'll examine daily accident data from Prince George's County to identify statistically unusual patterns that might lead to newsworthy stories. Z-scores provide a powerful statistical method to determine which days experienced an abnormal number of accidents compared to typical patterns.

A z-score tells us how many standard deviations a data point is from the mean. Generally:
- Z-scores between -2 and 2 represent typical values (about 95% of the data)
- Z-scores beyond ±2 indicate unusual values that might warrant investigation
- Z-scores beyond ±3 are very unusual (only about 0.3% of normally distributed data)

## Setup

Let's start by loading the necessary packages:

```{r load-packages}
library(tidyverse)  # For data manipulation and visualization
library(knitr)      # For nice tables
library(lubridate)  # For working with dates
```

### Task 1: Loading the Data (1 point)

Load the accident data from our CSV file at https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/daily_accidents.csv into a dataframe called `accidents`:

```{r load-data}
accidents <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/daily_accidents.csv")
```

## Data Exploration

Let's first understand our dataset structure:

```{r data-exploration}
# Get a summary of the data
summary(accidents)

# Check how many days we have in total
nrow(accidents)

# Check the date range
min(accidents$date)
max(accidents$date)
```

Now let's create a time series visualization of accidents:

```{r time-series-plot}
# Plot time series
ggplot(accidents, aes(x = date, y = total)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.1) +
  theme_minimal() +
  labs(title = "Daily Accidents in Prince George's County",
       x = "Date",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

### Task 2 (3 points)
Looking at the time series visualization above, identify at least two noticeable patterns or anomalies. What questions would you want to investigate further based on this initial view of the data? How might these visual patterns translate into potential story angles?

There was a period in what looks to be March-April 2020 when accident numbers were quite low compared to the general pattern of accidents between 2017 and 2025. I assume this was because COVID-19 quarantine began around this time, and people generally were driving less than normal. Something else that sticks out to me is the dramatic high number of car accidents one day near the end of 2021. This is already abnormal on its own, but what's even more interesting is that soon after that day, car accidents reached a low in the data set. As I journalist, I would investigate the causes behind these extreme numbers and also what they mean in relation to each other. How is it possible that the most and least number of daily car accidents in Prince George's County over a more than eight year span occurred so close together? The visual of this extreme opposition would make for an interesting story angle. 

The daily data is quite noisy, so let's also look at weekly patterns:

```{r weekly-patterns}
# Add day of week 
accidents <- accidents |>
  mutate(
    day_of_week = wday(date, label = TRUE),
    month = floor_date(date, "month"),
    year = year(date)
  )

# Check accident patterns by day of week
day_of_week_summary <- accidents |>
  group_by(day_of_week) |>
  summarize(
    avg_accidents = mean(total),
    median_accidents = median(total),
    max_accidents = max(total),
    min_accidents = min(total)
  )

# Display day of week summary
day_of_week_summary |> kable(digits = 2)

# Visualize by day of week
ggplot(accidents, aes(x = day_of_week, y = total)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Accident Distribution by Day of Week",
       x = "Day of Week",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

### Task 3 (2 points)
Based on the day-of-week analysis, which day(s) have the highest and lowest average accidents? Why might these patterns exist?

Friday and Saturday have the highest average accidents, while Monday and Wednesday have the lowest. Friday and Saturday's high averages might be due to drunk drivers. More people drink on weekend nights because they don't have work, meaning more people are by extension also getting behind the wheel while intoxicated, resulting in more car accidents. There could also just be more traffic on Fridays and Saturdays because people are taking trips/going new places on their days off. In contrast, Monday and Wednesday (and weekdays in general) probably have lower car accident averages because people are mostly just driving to work or school, routes they are used to driving. There are also likely far fewer drunk drivers on the road during the week. 

## Calculating Z-Scores

### Task 4 (2 points)

Now, using the accidents dataframe, add three columns using `mutate()`: mean_accidents, sd_accidents and z_score, based on the example from class.

```{r calculate-z-scores}
# Calculate z-scores for total accidents
accidents <- accidents |>
  mutate(
    mean_accidents = mean(total),
    sd_accidents = sd(total),
    z_score = (total - mean_accidents) / sd_accidents
  )

# Display the data with z-scores
accidents |>
  select(date, day_of_week, total, z_score) |>
  head(10) |>
  kable(digits = 2)
```

## Identifying Unusual Days

Let's identify days with unusually high or low accident numbers (|z-score| > 2):

```{r unusual-days}
# Find days with unusual accident numbers
unusual_days <- accidents |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

# Display number of unusual days
cat("Number of days with unusual accident counts:", nrow(unusual_days), 
    "out of", nrow(accidents), "total days (", 
    round(nrow(unusual_days)/nrow(accidents)*100, 2), "%)")

# Display the top 10 most unusual days
unusual_days |>
  head(10) |>
  select(date, day_of_week, total, z_score) |>
  kable(digits = 2)
```

### Task 5 (3 points)
Choose one of the top 10 unusual days and research what might have happened on that date. Look for weather events, holidays, major incidents, or other factors that could explain the unusual number of accidents. Briefly describe your findings and how you would incorporate this context into a story.

I chose to research October 29, 2021, which was the worst recorded day for car accidents in Prince George's County from 2017-2025. I discovered that there was a major storm in Maryland that day, so bad that Somerset and Dorchester counties experienced the worst Chesapeake Bay tidal flooding in the state for 50 years, according to a press release from Maryland's Voluntary Organizations Active in Disaster. Even though Prince George's County is about 100-150 miles away from those counties and sits far inland from the Bay, I'm sure the same storm that caused the extreme flooding in that area also hit Prince George's, causing an unusual amount of car accidents. I would definitely mention this weather event in my story as context for why there were so many car accidents in the county on October 29, 2021, a day that otherwise was not out of the ordinary for any reason.  

Let's visualize these unusual days:

```{r plot-unusual-days}
# Highlight unusual days in the time series
ggplot(accidents, aes(x = date, y = total, color = abs(z_score) > 2)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("gray", "red"), 
                     name = "Unusual Value",
                     labels = c("Normal", "Unusual")) +
  theme_minimal() +
  labs(title = "Daily Accidents with Unusual Days Highlighted",
       subtitle = "Unusual days have z-scores with absolute values > 2",
       x = "Date",
       y = "Number of Accidents",
       caption = "Source: Prince George's County Accident Data")
```

## Day of Week Patterns

### Task 6 (3 points)

Let's calculate z-scores that account for day of week patterns based on the same pattern as above:

```{r day-of-week-z-scores}
# Calculate day-of-week adjusted z-scores
accidents <- accidents |>
  group_by(day_of_week) |>
  mutate(
    dow_mean = mean(total),
    dow_sd = sd(total),
    dow_z_score = (total - dow_mean) / dow_sd
  ) |>
  ungroup()

# Find days with unusual day-of-week z-scores
unusual_dow <- accidents |>
  filter(abs(dow_z_score) > 2) |>
  arrange(desc(abs(dow_z_score)))

# Display the top unusual days adjusted for day of week
unusual_dow |>
  head(10) |>
  select(date, day_of_week, total, dow_mean, dow_z_score) |>
  kable(digits = 2)
```

### Task 7 (3 points)
Compare the days that have extreme z-scores overall versus those with extreme day-of-week z-scores. How do these lists differ? What journalistic insights can you draw from comparing these two different approaches to measuring "unusual" days?

The list of days that have extreme z-scores overall is mostly composed of Fridays and Saturdays, as those are the days of the week that see the highest number of accidents. On the other hand, the list of days with extreme day-of-the-week z-scores has a lot more weekdays. The two lists share some of the same days of the week, those that are the most extreme outliers overall such as Friday, Oct. 29, 2021, and Monday, Nov. 29, 2021, but the day-of-the-week list includes days such as Wednesday, May 5, 2021 and Tuesday, Dec. 10, 2024 because it takes day of the week into account when calculating how unusual a day's number of car accidents is. For example, Tuesday, Dec. 10, 2024 may not be one of the most extreme outliers overall, but compared to all other Tuesdays, it is quite unusual. As a journalist, I think it would be interesting to examine these abnormal weekdays and investigate what led to their either very high or very low number of car accidents.   

Let's see how many unusual days we have for each day of the week:

```{r unusual-by-dow}
# Count unusual days by day of week
unusual_count_by_dow <- accidents |>
  filter(abs(dow_z_score) > 2) |>
  count(day_of_week) |>
  arrange(desc(n))

# Get total count by day of week for comparison
total_by_dow <- accidents |>
  count(day_of_week)

# Join to calculate percentages
unusual_rates <- unusual_count_by_dow |>
  left_join(total_by_dow, by = "day_of_week") |>
  rename(unusual_days = n.x, total_days = n.y) |>
  mutate(
    pct_unusual = (unusual_days / total_days) * 100
  )

# Display results
unusual_rates |> kable(digits = 2)
```

## Monthly Patterns

Let's examine if certain months have more unusual days:

```{r monthly-patterns}
# Extract month from date for seasonal analysis
accidents <- accidents |>
  mutate(month_of_year = month(date, label = TRUE))

# Calculate average accidents by month
monthly_pattern <- accidents |>
  group_by(month_of_year) |>
  summarize(
    avg_accidents = mean(total),
    median_accidents = median(total)
  )

# Plot monthly pattern
ggplot(monthly_pattern, aes(x = month_of_year, y = avg_accidents, group = 1)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Average Daily Accidents by Month",
       x = "Month",
       y = "Average Number of Accidents",
       caption = "Source: Prince George's County Accident Data")

# Calculate month-adjusted z-scores
accidents <- accidents |>
  group_by(month_of_year) |>
  mutate(
    monthly_mean = mean(total),
    monthly_sd = sd(total),
    monthly_z_score = (total - monthly_mean) / monthly_sd
  ) |>
  ungroup()

# Find days with unusual monthly z-scores
unusual_monthly <- accidents |>
  filter(abs(monthly_z_score) > 2) |>
  arrange(desc(abs(monthly_z_score)))

# Display top days that are unusual for their month
unusual_monthly |>
  head(10) |>
  select(date, month_of_year, day_of_week, total, monthly_mean, monthly_z_score) |>
  kable(digits = 2)
```

### Task 8 (2 points)
Based on the monthly pattern analysis, which months have the highest and lowest average accidents? Develop a hypothesis about why these seasonal patterns exist.

October has the highest average accidents and January has the lowest. I think October might have the highest average because that is when we turn back the clocks and it begins to get darker out much earlier in the evening, when people might still be driving home from work. Extreme weather events (snow, rainstorms) also tend to pick up in the fall, which could explain why there are so many car accidents in October. I think January might have the lowest average because people don't tend to travel as much during that month. They are probably spent from travelling during the holiday season and more inclined to stay home. They could also be trying to save money at the start of the new year and therefore refrain from driving and wasting gas.  

## Identifying Extreme Outliers

Let's look specifically at days with extremely unusual accident counts (z-scores beyond ±3):

```{r extreme-outliers}
# Find extreme outliers
extreme_outliers <- accidents |>
  filter(abs(z_score) > 3) |>
  arrange(desc(abs(z_score)))

# Display extreme outliers
if(nrow(extreme_outliers) > 0) {
  extreme_outliers |>
    select(date, day_of_week, total, z_score) |>
    kable(digits = 2)
} else {
  cat("No days with z-scores beyond ±3 were found in the dataset.")
}
```

### Task 9 (2 points)
For the extreme outliers identified (z-scores beyond ±3), what do they have in common?

Most of the days are Fridays or Saturdays, and there is also one Sunday and one Monday. All of the extreme days saw either more than 45 car accidents or less than four car accidents.  

## Your Best Story Pitch

### Task 10 (5 points)
Based on all the analyses performed, craft your best story pitch, including a headline and lede. Be sure to include what kinds of sources/people you'd talk to for the story.

Idea: Why was there only one recorded car accident in Prince George's County on Monday, November 29, 2021? What happened on that day? Was there actually only one accident, or were there more that just went unrecorded for some reason? This story would dive into the past and examine what went on that day that might explain its historic low car-accident numbers. It would also provide details about the one car accident that did happen that day: who was involved, what caused it, how much damage was done, etc.

Headline: Over an 8-year period, this was calmest day of driving in Prince George's County

Lede: From February 2017 to April 2025, the average number of daily car accidents recorded in Prince George's County was just about 25, according county accident data. Nothing crazy, but certainly enough to keep police officers busy most days each year. Except for one unusual day toward the end November 2021, when just a singular car accident was recorded. Here's what happened to account for the safe roads. 

Sources: Prince George's County police department, accident records, those involved in the one recorded accident, weather specialists
